---
title: "ENVS193DS Final"
author: "Callie Baker"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---

# GitHub repo
Access my repo [here](https://github.com/c-baker3/ENVS-193DS_spring-2026_final).

# Set up

```{r}
#| label: packages and data

# read in packages
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames

# read in data
nests <- read_csv(here("data", "nest_data_final.csv"))

```


# Problem 1. Research writing

## a. Transparent statistical methods


## b. Suggestions for rewriting


## c. Further analysis


# Problem 2. Data analysis

*DO NOT include any code or output from your data exploration

## a. Clean and wrangle

```{r}
#| label: cleaning

# cleaned data
nests_clean <- nests |> 
  clean_names()



```

## b. Response variable


## c. Purpose of study


## d. Table of models


## e. Run the models


## f. Select the best model


## g. Check the diagnostics


## h. Visualize the model predictions 


## i. Write the caption for your figure


## j. Calculate model predictions


## k. Interpret your results


# Problem 3. Affective and exploratory visualizations

## a. Comparising visualizations


## b. Designing an analysis


## c. Check any assumptions and run your analysis


## d. Create a visualization


## e. Write a caption


## f. Write about your results


## g. Sharing your affective visualization





